Parsimonious Representation of Nonlinear Dynamical Systems using Diffusion Maps: A Chemotaxis Case Study

### 2015-02-25 chemotaxis/paper/draft1.tex
small edits to paper
updated figures for chemotaxis

\section{Background on Manifold Learning and Some Relevant Issues}

@@

\subsection{Identifying the Informative Eigenvectors using Local Linear Regression}

@@

@@

@@

We will demonstrate the local linear regression algorithm for identifying informative eigenvectors on two synthetic data sets. 
%
Both of these data sets are nonlinear manifolds which provide examples of the flexibility of diffusion maps. 
%
However, these data sets will also illustrate the issues of harmonics.

@@

Our first illustrative example will consist of two different swiss roll data sets.
%
The data are
\begin{equation}
\begin{aligned}
z_1 =& \theta \cos \theta \\
z_2 =& \theta \sin \theta \\
z_3 =& h t
\end{aligned}
\end{equation}
%
where $\theta$ and $t$ are random variables, sampled such that the data is uniformly distributed along the swiss roll. 
%
The height of the first swiss roll is $h = 40$, while the height of the second is $h = 20$. 
%
Each data set consists of $m=1500$ points, shown shown in Figure~\ref{fig:swiss_rolls}(a)~and~(b).
%
We expect to recover the eigenvector which parameterizes the height farther down in the spectrum for the second data set relative to the first, as this second data set is shorter.

@@

For the second example, we consider a torus defined by
%
\begin{equation}
\begin{aligned}
z_1 =& (r_1 + r_2 \cos \theta_2 ) \cos \theta_1 \\
z_2 =& (r_1 + r_2 \cos \theta_2 ) \sin \theta_1 \\
z_3 =& r_2 \sin \theta_2
\end{aligned}
\end{equation}
%
where $r_1 > r_2$ are the outer and inner radii, respectively, and $\theta_1, \theta_2 \sim Unif(0, 2 \pi)$. 
%
Changing the ratio of $r_1$ to $r_2$ changes the ratio of the eigenvalues corresponding to non-harmonic eigenvectors.

@@

\begin{figure}[t]
%
\begin{subfigure}{\textwidth}
\includegraphics[width=2.5in]{torus1}
\includegraphics[height=1.5in]{torus1_evals}
\caption{}
\end{subfigure}
%
\begin{subfigure}{\textwidth}
\includegraphics[width=2.5in]{torus2}
\includegraphics[height=1.5in]{torus2_evals}\\
\caption{}
\end{subfigure}
%
\begin{subfigure}{\textwidth}
\includegraphics[width=2.5in]{torus3}
\includegraphics[height=1.5in]{torus3_evals}
\caption{}
\end{subfigure}
%
\caption{Torus data sets (left) and corresponding diffusion maps eigenvalues (right). (a) $r_1 = 3$, $r_2 = 1$. (b) $r_1 = 5$, $r_2 = 1$. (c) $r_1 = 10$, $r_2 = 1$. In all three data sets, the first two eigenvalues/eigenvectors correspond to $\sin \theta_1$ and $\cos \theta_1$. The second pair of non-harmonic eigenvectors (corresponding to $\sin \theta_2$ and $\cos \theta_2$) are captured by components 7 and 8, 11 and 12, and 15 and 16, respectively.}
%
\label{fig:torus}
%
\end{figure}

@@

To detect this transition in the data, we look at the eigenvalues corresponding to the two independent eigenevctors. 
%
Let $\mu_{i_1} \ge \mu_{i_2}$ denote these two eigenvalues. 
%
According to the relationship in \eqref{eq:evals} and \eqref{eq:evals_relationship}, we plot  
\begin{equation}
 \frac{\log \mu_{i_1}}{\log \mu_{i_2}} ;
\end{equation}
when this ratio becomes small, the data are effectively one-dimensional, as the second dimension is very small compared with the first. 
%
Figure~\ref{fig:chemotaxis_simulations_harmonics} shows the eigenvalue spectra for three different sets of simulations. 
%
It now becomes clear that looking at the diffusion maps coordinates {\em modulo} higher harmonics is essential for extracting an informative parameterization of the data, as well as characterizing the dimensionality of the data. 
%
From Figures~\ref{fig:chemotaxis_simulations_harmonics}(b-c), we can see that the eigenvalue spectrum does not exhibit any large spectral gaps.
%
However, there is a significant gap between the eigenvalues which correspond to new directions in the data, and the corresponding eigenvectors provide us with a two-dimensional parameterization of the data which is consistent with the expected macroscopic dynamics.

>>>>

\section{Manifold Learning and Some Relevant Issues}

@@

\begin{figure}[t]
\centering
\begin{subfigure}{0.3\textwidth}
\includegraphics[height=3cm]{strip_data_L2}
\includegraphics[height=3.5cm]{strip_spectrum_L2}
\caption{}
\end{subfigure}
%
\begin{subfigure}{0.3\textwidth}
\includegraphics[height=3cm]{strip_data_L4}
\includegraphics[height=3.5cm]{strip_spectrum_L4}
\caption{}
\end{subfigure}
%
\begin{subfigure}{0.35\textwidth}
\includegraphics[height=3cm]{strip_data_L8}
\includegraphics[height=3.5cm]{strip_spectrum_L8}
\caption{}
\end{subfigure}
%
\caption{Data sets (top) and eigenvalue spectra (bottom) for strips with (a) $L_1 = 2$, $L_2 = 1$, (b) $L_1 = 4$, $L_2 = 1$, (c) $L_1 = 8$, $L_2 = 1$. The empirical eigenvalues are plotted in blue, and the analytical eigenvalues are plotted in red. From the eigenvalues which are identified as important/non-harmonics, the estimated length ratio is (a) 2.0, (b) 4.4, (c) 8.8.  }
\label{fig:strip_compare_analytic}
\end{figure}

\section{Identifying the Informative Eigenvectors }

@@

\subsection{Algorithm: Local Linear Regression}

@@

\subsection{Calculating the relative lengths of the manifold}

Once we have determined which eigenvectors parameterize independent directions along the manifold, we can use the corresponding eigenvalues to calculate the relative lengths along these directions. 
%
Let $\mu_{i_1}, \mu_{i_2}, \dots$ denote the eigenvalues which correspond to eigenvectors which are {\em not} higher harmonics. 
%
From \eqref{eq:evals} and \eqref{eq:evals_relationship}, we propose to use
\begin{equation} \label{eq:est_lengths}
L_j \propto \frac{1}{\sqrt{-\log \mu_{i_j}}}
\end{equation}
where $L_1, L_2, \dots$ are the relative lengths along the principal directions of the manifold. 
%
We can then use this measure to determine how many components are required to retain most of the information within the data set.

@@

We will demonstrate the local linear regression algorithm for identifying informative eigenvectors on three synthetic data sets. 
%
The first data set is the strip where the solution is known analytically, and the second and third data sets are nonlinear manifolds which provide examples of the flexibility of diffusion maps. 
%
However, these data sets will also illustrate the issues of harmonics.

\subsubsection{Strip}

Our first illustrative example consists of three different strip data sets. 
%
Each of these data sets contains $m=1500$ data points uniformly distributed along the strip. 
%
Figure~\ref{fig:strip_compare_analytic} shows the data sets and corresponding eigenspectra.
%
The diffusion maps eigenvalues are consistent with the known analytic eigenvalues of the Laplacian, and using \eqref{eq:est_lengths}, we can accurately estimate the relative lengths of the two principal directions in each data set.

@@

Our second illustrative example will consist of two different swiss roll data sets.
%
The data are
\begin{equation}
\begin{aligned}
z_1 =& \theta \cos \theta \\
z_2 =& \theta \sin \theta \\
z_3 =& h t
\end{aligned}
\end{equation}
%
where $\theta$ is uniformly sampled along the arclength of the spiral, and $t \sim Unif(0,1)$. 
%
The height of the first swiss roll is $h = 40$, while the height of the second is $h = 20$. 
%
Each data set consists of $m=1500$ points, shown shown in Figure~\ref{fig:swiss_rolls}(a)~and~(b).
%
We expect to recover the eigenvector which parameterizes the height farther down in the spectrum for the second data set relative to the first, as this second data set is shorter.

@@

For the third example, we consider a torus defined by
%
\begin{equation}
\begin{aligned}
z_1 =& (r_1 + r_2 \cos \theta_2 ) \cos \theta_1 \\
z_2 =& (r_1 + r_2 \cos \theta_2 ) \sin \theta_1 \\
z_3 =& r_2 \sin \theta_2
\end{aligned}
\end{equation}
%
where $r_1 > r_2$ are the outer and inner radii, respectively, and $\theta_1, \theta_2 \sim Unif(0, 2 \pi)$. 
%
Changing the ratio of $r_1$ to $r_2$ changes the ratio of the eigenvalues corresponding to non-harmonic eigenvectors.

@@

\begin{figure}[t]
%
\begin{subfigure}{0.3\textwidth}
\includegraphics[height=0.75in]{torus1}
\includegraphics[height=1in]{torus1_evals}
\caption{}
\end{subfigure}
%
\begin{subfigure}{0.32\textwidth}
\includegraphics[height=0.75in]{torus2}
\includegraphics[height=1in]{torus2_evals}
\caption{}
\end{subfigure}
%
\begin{subfigure}{0.35\textwidth}
\includegraphics[height=0.75in]{torus3}
\includegraphics[height=1in]{torus3_evals}
\caption{}
\end{subfigure}
%
\caption{Torus data sets (left) and corresponding diffusion maps eigenvalues (right). (a) $r_1 = 3$, $r_2 = 1$. (b) $r_1 = 5$, $r_2 = 1$. (c) $r_1 = 10$, $r_2 = 1$. In all three data sets, the first two eigenvalues/eigenvectors correspond to $\sin \theta_1$ and $\cos \theta_1$. The second pair of non-harmonic eigenvectors (corresponding to $\sin \theta_2$ and $\cos \theta_2$) are captured by components 7 and 8, 11 and 12, and 15 and 16, respectively.}
%
\label{fig:torus}
%
\end{figure}

@@

To detect this transition in the data, we look at the eigenvalues corresponding to the two independent eigenevctors. 
%
Let $\mu_{i_1} \ge \mu_{i_2}$ denote these two eigenvalues. 
%
According to \eqref{eq:est_lengths}, we plot  
\begin{equation}
 \sqrt{\frac{\log \mu_{i_1}}{\log \mu_{i_2}}} ;
\end{equation}
when this ratio becomes small, the data are effectively one-dimensional, as the second dimension is very small compared with the first. 
%
Figure~\ref{fig:chemotaxis_simulations_harmonics} shows the eigenvalue spectra for three different sets of simulations. 
%
It now becomes clear that looking at the diffusion maps coordinates {\em modulo} higher harmonics is essential for extracting an informative parameterization of the data, as well as characterizing the dimensionality of the data. 
%
From Figures~\ref{fig:chemotaxis_simulations_harmonics}(b-c), we can see that the eigenvalue spectrum does not exhibit any large spectral gaps.
%
However, there is a significant gap between the eigenvalues which correspond to new directions in the data, and the corresponding eigenvectors provide us with a two-dimensional parameterization of the data which is consistent with the expected macroscopic dynamics.
